---
title: "paramaters"
author: "Thijmen"
output: bookdown::html_document2
params:
  country:
    label: "Country"
    input: select
    choices: [Austria, Belgium, Bulgaria, Croatia, Cyprus, Czechia, Denmark, Estonia, Finland, France, Germany, Greece, Hungary, Iceland, Ireland, Italy, Latvia, Liechtenstein, Lithuania, Luxembourg, Malta, Netherlands, Norway, Poland, Portugal, Romania, Slovakia, Slovenia, Spain, Sweden]
    value: Netherlands
  year:
    label: "Year"
    min: 2020
    max: 2022
    step: 1
    input: slider
    sep: ""
    value: 2020
  period_start:
    label: "Period start" 
    min: 1
    max: 12
    step: 1
    input: slider
    sep: ""
    value: 1
  period_end:
    label: "Period end"
    min: 1
    max: 12
    step: 1
    input: slider
    sep: ""
    value: 12
---

## Creating paramaters for different data inputs

To show my ability to use paramaters I will be using data from [the ECDC](https://www.ecdc.europa.eu/en/publications-data/data-daily-new-cases-covid-19-eueea-country). the data is available in this repository under "data/COVID_cases_31_05_2022"

```{r setup, include=FALSE}
library(plotly)
library(tidyverse)
```


```{r creating the graphs, warning=FALSE, error=FALSE}

# loading in data
cases <- read.csv("data/COVID_cases_31_05_2022.csv")

# filtering the params used
cases_filtered <- cases %>% dplyr::filter(countriesAndTerritories == params$country, year == params$year, month >= params$period_start, month <= params$period_end) 

# telling R the dateRep column is a date
cases_filtered$dateRep <- as.Date(cases_filtered$dateRep, format = "%d/%m/%Y")

# making a graph for cases
cases_graph <- cases_filtered %>%
  ggplot(aes(x = dateRep, y = cases)) +
  geom_point(size = .5) +
  geom_line() +
  labs(title = paste("Covid related cases from month", params$period_start, "to", params$period_end, "in", params$year, "for", params$country),
       x = "Month",
       y = "Covid related cases") +
  theme_classic()

ggplotly(cases_graph)

# making a graph for deaths
deaths_graph <- cases_filtered %>%
  ggplot(aes(x = dateRep, y = deaths)) +
  geom_point(size = .5) +
  geom_line() +
  labs(title = paste("Covid related deaths from month", params$period_start, "to", params$period_end, "in", params$year, "for", params$country),
       x = "Month",
       y = "Covid related deaths") +
  theme_classic()

ggplotly(deaths_graph)
```

If you want to recreate these graphs with different parameters, clone this repository and use put this command in the console (with your own params ofcourse): `bookdown::render_book(params = list(country = "Netherlands", year = 2021, period_start = 5, period_end = 10))`
